---
title: "EBIO 3080 Week 1 `R` Tutorial"
author: "Silas Tittes"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: beamer_presentation
---

```{r, cache=FALSE, echo=FALSE, message=FALSE, error=FALSE}
library(knitr)
library(formatR)
opts_chunk$set(comment = "", warning = FALSE, 
               message = FALSE, 
               echo = TRUE, tidy = FALSE, size="small",
               tidy.opts=list(width.cutoff=75))
```


## Goals
>- Basics

> - Navigating your computer like a pro (sort of) -- file paths with getwd(), setwd(), and list.files()

> - Read data from a text file or csv into `R` -- read.csv() or read.table()

> - Working with data frames -- the $ operator

> - Basic math and summary statistics on vectors

> - Plotting data -- plot() and hist()


##Basics
> - Write all commands in a scipt and save with .R extension. Sparingly use the console directly.
> - To run a command, use the  "Control" + "Enter" keys, line does not need to be highlighted,
but multiple lines can be highlighted if you want to run multiple lines.
> - Assignment operator to create a new variable/object "<-" or "=". 
> - Hotkey for making "<-" is "Alt" + "-" keys.
> - Use "Control" + "a" to highlight all code, then "Control" + "Enter" to run all code.
> - Use the "tab" key and `R` will often complete or predict the remaining text. *Very useful*!

 

## Navigating your computer like a pro
To be effecient in `R` we need to know where we are and where we want to go within our computer.
```{r}
#try using the tab button for 
#suggestions on how to finish your path

#get working directory
getwd()
#set working directory
setwd("/home/silastittes/Desktop/")
getwd()
setwd("~/Documents/TAing/Fall16_Evolution/week1/")
```


## Navigating your computer like a pro
Make sure your files are where you think they are!
```{r}
list.files()
```


##Read data from a text file or csv into R
> - Use read.csv() to read in a csv file.
> - Use read.table() to read in a txt file.
> - Once you know where yourself and the data files you need are, read data into R.
> - Expect data to almost always be in csv format for this class.
> - Always save data to an object, with a meaningful name.
> - Let `R` know the first row of the data set are names for the columns using "header = TRUE"
```{r}
lizard <- read.csv("horned_lizards.csv", header = TRUE)
```


##Working with data frames
Take a quick look at the data with head(), which prints
the first six rows be default.
```{r}
head(lizard)
```


##Working with data frames
Work with individual columns of the data frame with the $ operator.
```{r}
head(lizard$Survival)
head(lizard$squamosalHornLength)
```


##Working with data frames
Get rows or columns of the data frame with indexing brackets
```{r}
#first row all columns
lizard[1,]

#second row first column
lizard[2,1]

#last row second column
lizard[nrow(lizard),2]

```


##Basic math and summary statistics on vectors
- `R` likes vectors -- 1 x n dimnsional sets of values all of the same type.
- Use the c() function to construct a new vectors.
```{r}
#numeric vector
n <- c(1, 2, 3)
n
is.vector(n) #if a vector, will print TRUE

#character vector
ch <- c("a", "b", "c")
ch
is.vector(ch)
```


##Basic math and summary statistics on vectors
Math operators are pretty straight forward.
```{r}
n + n
n * n
n^2
sqrt(n)
log(n)
log(n, base = 2)
```


##Basic math and summary statistics on vectors
`R` thinks of columns of a data frame as a vector, so rules for vectors usually apply to columns of a data frame.   
```{r}
#don't use head!
head(lizard$squamosalHornLength)
head(lizard$squamosalHornLength * 2)
```


##Basic math and summary statistics on vectors
`R` has an immense statistical library
```{r}
mean(lizard$squamosalHornLength) #DOH!
mean(lizard$squamosalHornLength, na.rm = TRUE)
median(lizard$squamosalHornLength, na.rm = TRUE)
var(lizard$squamosalHornLength, na.rm = TRUE)
sd(lizard$squamosalHornLength, na.rm = TRUE)
summary(lizard$squamosalHornLength, na.rm = TRUE)
```

##Plotting data
`R` also has some incredible graphics utilities
```{r, eval=F}
boxplot(lizard$squamosalHornLength ~ lizard$Survival, 
        ylab = "Horn length",
        col=c("grey40", "purple"),
        cex.lab=1.4)

legend("topleft", c("Dead", "Alive"), 
       pch=19, col=c("grey40", "purple"), bty="n")
```

##Plotting data
```{r, echo=F}
boxplot(lizard$squamosalHornLength ~ lizard$Survival, 
        ylab = "Horn length",
        col=c("grey40", "purple"),
        cex.lab=1.4)

legend("topleft", c("Dead", "Alive"), 
       pch=19, col=c("grey40", "purple"), bty="n")
```


##Plotting data
`R` also has some incredible graphics utilities
```{r, eval=FALSE}
dead <- lizard[lizard$Survival == "killed", ]
alive <- lizard[lizard$Survival != "killed", ]

rng <- range( c(dead$squamosalHornLength, alive$squamosalHornLength), na.rm=T)

hist(alive$squamosalHornLength,
     xlab = "Horn Length",
     cex.lab=1.4, 
     cex=0.5, main = "", col = "purple")

hist(dead$squamosalHornLength, xlim=rng,
     xlab = "Horn Length",
     cex.lab=1.4, 
     cex=0.5, main = "", col = "grey40", add=T)


legend("topleft", c("Dead", "Alive"), 
       pch=19, col=c("grey40", "purple"), bty="n")
```

##Plotting data
```{r, echo=FALSE}
dead <- lizard[lizard$Survival == "killed", ]
alive <- lizard[lizard$Survival != "killed", ]

rng <- range( c(dead$squamosalHornLength, alive$squamosalHornLength), na.rm=T)

hist(alive$squamosalHornLength,
     xlab = "Horn Length",
     cex.lab=1.4, 
     cex=0.5, main = "", col = "purple")

hist(dead$squamosalHornLength, xlim=rng,
     xlab = "Horn Length",
     cex.lab=1.4, 
     cex=0.5, main = "", col = "grey40", add=T)


legend("topleft", c("Dead", "Alive"), 
       pch=19, col=c("grey40", "purple"), bty="n")
```


##Plotting data
`R` also has some incredible graphics utilities
```{r, eval=FALSE}
plot(sort(lizard$squamosalHornLength),
     ylab="Sorted horn length", 
     pch=19, cex.lab=1.4)
```


##Plotting data
```{r, echo=FALSE}
plot(sort(lizard$squamosalHornLength),
     ylab="Sorted horn length", 
     pch=19, cex.lab=1.4)
```


##Plotting data
`R` also has some incredible graphics utilities
```{r, eval=FALSE}
surv <- as.numeric(lizard$Survival)-1
colz <- ifelse(surv == 1, "grey40", "purple")
plot(lizard$squamosalHornLength, jitter(surv, 0.1),
     xlab="Sorted horn length", ylab = "Survival",
     pch=19, col = colz, cex.lab=1.4)


legend("topleft", c("Dead", "Alive"), 
       pch=19, col=c("grey40", "purple"), bty="n")
```

##Plotting data
```{r, echo=FALSE}
surv <- as.numeric(lizard$Survival)-1
colz <- ifelse(surv == 1, "grey40", "purple")
plot(lizard$squamosalHornLength, jitter(surv, 0.1),
     xlab="Sorted horn length", ylab = "Survival",
     pch=19, col = colz, cex.lab=1.4)


legend("topleft", c("Dead", "Alive"), 
       pch=19, col=c("grey40", "purple"), bty="n")
```
